---
title: "README"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Install required packages if not already installed
if (!require("httr")) install.packages("httr", dependencies = TRUE)

if (!require("kableExtra")) install.packages("kableExtra", dependencies = TRUE)
if (!require("knitr")) install.packages("knitr", dependencies = TRUE)
library(knitr)

if (!require("psych")) install.packages("psych", dependencies = TRUE)
library(psych)

# Set Kaggle credentials (Ensure your Kaggle API key is in ~/.kaggle/kaggle.json)
Sys.setenv(KAGGLE_CONFIG_DIR = "~/.kaggle")

library(fBasics)

```

## Loading Data
```{r}
# Define the URL of the dataset
url <- "https://www.kaggle.com/api/v1/datasets/download/fratzcan/usa-house-prices"

# Define the destination file path 
USA_housing_dataset <- "usa-house-prices.zip"

# Extract the dataset
unzip(USA_housing_dataset, exdir = "usa-house-prices")

# List extracted files 
list.files("usa-house-prices")

# Import dataset 
housing_data <- read.csv ("USA Housing Dataset.csv", stringsAsFactors = FALSE)
```
```{r, include=FALSE}
#getwd()
```

```{r, echo=FALSE}
#Getting summary statistics of the dataset

#head(housing_data)
#summary(housing_data)
#str(housing_data)

#install.packages("DT")
#library(DT)
#datatable(housing_data)

#View(housing_data)

#List column names
#names(housing_data)
```

## PART 1: Understanding the Data

### When was the data acquired?

The data was acquired from May 2, 2014 to July 10, 2014. The data was uploaded to Kaggle and was first accessed for this analysis on February 12, 2025.
```{r, eval=TRUE, include=FALSE}
mind <- min(housing_data$date, na.rm = TRUE)
maxd <- max(housing_data$date, na.rm = TRUE)

#ANSWER: > min(housing_data$date, na.rm = TRUE)
#[1] "2014-05-02 00:00:00"
#> max(housing_data$date, na.rm = TRUE)
#[1] "2014-07-10 00:00:00"
```


### Where was the data acquired? (geographical location)
```{r, include=FALSE}
table(housing_data$city)
city_table <- as.data.frame(table(housing_data$city))
colnames(city_table) <- c("City", "Frequency")
where <- city_table
#View(city_table)
```
The data consists of housing prices from various cities in Washington, US. The cities are shown in the frequency table below:
`r knitr::kable(where)`

### How was the data acquired?

This data was acquired from Kaggle. The page on Kaggle does not contain any information about the source. There are no links to original publications or projects from which the data was derived.

```{r, include=FALSE}
# Create a data frame with attribute names and descriptions
attributes_table <- data.frame(
  Attribute = c("date", "price", "bedrooms", "bathrooms", "sqft_living", "sqft_lot", "floors", "waterfront", "view", "condition",  "sqft_above", "sqft_basement", "yr_built", "yr_renovated", "street", "city", "statezip", "country"),
 
  Description = c("Date when the house was sold",
                  "Sale price of the house",
                  "Number of bedrooms",
                  "Number of bathrooms",
                  "Living area in square feet",
                  "Lot area in square feet",
                  "Number of floors",
                  "Whether the house is waterfront (1: Yes, 0: No)",
                  "Number of times the house has been viewed",
                  "Condition of the house (1-5 scale)",
                  "Square footage of the house above ground",
                  "Square footage of the basement",
                  "Year the house was built",
                  "Year of renovation (0 if never renovated)",
                  "Street address of house", 
                  "City of the house location",
                  "Zip code of the house location",
                  "Country of the house location")
)

# Print the table
print(attributes_table)
```

```{r, include=FALSE}
str(housing_data, list.len = ncol(housing_data)) #Check structure of data
```

```{r,include=FALSE}
#View(attributes_table)
# Check data type of each attribute
print(typeof(housing_data$date))
print(typeof(housing_data$price))
print(typeof(housing_data$bedrooms))
print(typeof(housing_data$bathrooms))
print(typeof(housing_data$sqft_living))
print(typeof(housing_data$sqft_lot))
print(typeof(housing_data$floors))
print(typeof(housing_data$waterfront))
print(typeof(housing_data$view))
print(typeof(housing_data$condition))
print(typeof(housing_data$sqft_above))
print(typeof(housing_data$sqft_basement))
print(typeof(housing_data$yr_built))
print(typeof(housing_data$yr_renovated))
print(typeof(housing_data$street))
print(typeof(housing_data$city))
print(typeof(housing_data$statezip))
print(typeof(housing_data$country))
```
### What type of data do these attributes contain? (Categorical, numerical, ordinal, etc)

The dataset consists of 18 different attributes. The attributes of the data and the data type of each attribute are shown in the table below:

```{r, include=FALSE}
# Add a column corresponding to data type to the attributes table
attributes_table$Type = c("character", "double", "double", "double", "integer", "integer", "double", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "character", "character", "character", "character")

# Generate a well-formatted table
styled_table <- kable(attributes_table, format = "html", caption = "📌 USA Housing Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE, color = "black") %>%
  column_spec(2, italic = TRUE, color = "blue") %>%
  column_spec(3, width = "40em") %>%
  row_spec(0, bold = TRUE, color = "white", background = "black")

#View(attributes_table)
```

```{r, echo=FALSE}
# Display the table
styled_table
```

## PART 2: Data Summary & Initial Insights 
```{r,evaluate=TRUE, include=FALSE}
#Summary statistics for each attribute
#describe(housing_data)
#summary(housing_data$price)
#describe(housing_data$price)

# Select only numeric columns
numeric_data <- housing_data[sapply(housing_data, is.numeric)]
```

```{r}
# Create a summary table, ensuring NAs are handled correctly
summary_table <- data.frame(
  Attribute = colnames(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) median(x, na.rm = TRUE)),
  SD = sapply(numeric_data, function(x) sd(x, na.rm = TRUE))
)

```

Here is a table that shows the code that corresponds with each attribute. These codes are useful for calculating summary statistics.

```{r}
Attribute <- colnames(housing_data)  # Extract column names
Code <- seq_along(Attribute)  # Create a sequence for the codes

# Create a proper data frame
df <- data.frame(Code, Attribute, stringsAsFactors = FALSE)

# Generate a well-formatted table
df %>%
  kable(caption = "Attribute Codes for Housing Data", longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)
```

```{r}
colnames (housing_data) = Code
housing_data3 = housing_data[-c(1:18)]
housing_data2 = housing_data[-c(1, 15:18)]
bstats = basicStats (housing_data2)[c("Mean", "Stdev", "Median", "Minimum", "Maximum",
"NAs"), ]
```

```{r, eval=TRUE, include=FALSE}
m<-matrix(1:ncol(bstats), 3)
for (i in 1:ncol(m)) {
cat(kable(bstats[, m[, i]], "latex", booktabs = TRUE, digits = 2, ), "\\newline")
}
```

```{r, echo=FALSE}
colnames(housing_data) <- Code
housing_data3 = housing_data[-c(1:18)]
housing_data2 = housing_data[-c(1, 15:18)]
bstats = basicStats (housing_data2)[c("Mean", "Stdev", "Median", "Minimum", "Maximum",
"NAs"), ]
```

```{r, eval=TRUE, include=FALSE}
m<-matrix(1:ncol(bstats), 3)
for (i in 1:ncol(m)) {
cat(kable(bstats[, m[, i]], "latex", booktabs = TRUE, digits = 2, ), "\\newline")
}
```

```{r, include=FALSE}
library(tidyr)
library(dplyr)
library(kableExtra)

summary_stats <- bstats %>%
  summarise(across(everything(), list(
    Min = ~min(.x, na.rm = TRUE),
    Q1 = ~quantile(.x, 0.25, na.rm = TRUE),
    Median = ~median(.x, na.rm = TRUE),
    Mean = ~mean(.x, na.rm = TRUE),
    Q3 = ~quantile(.x, 0.75, na.rm = TRUE),
    Max = ~max(.x, na.rm = TRUE),
    SD = ~sd(.x, na.rm = TRUE)
  ), .names = "{.col}_{.fn}"))

summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

print(summary_long)
```

```{r}
summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))
```

The table created below shows summary statistics for the attributes:
```{r}
summary_long %>%
  kable(caption = "Summary Statistics for Housing Data", longtable = TRUE, align = "c") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%  # Bold column names (first row)
  row_spec(1:nrow(summary_long), font_size = 12) %>%  # Adjust font size for summary rows
  column_spec(1, bold = TRUE)   # Bold the first column (Statistic names)
```

### Identify any missing (NA) or empty values and discuss potential strategies to handle them.

When calculating summary statistics and creating our summary table, we separated numeric data and non-numeric data. There are 5 attributes of the character data type, so these attributes had empty values. We did not include these categorical attributes in our analysis.  

### If applicable, include visualizations such as histograms, box plots, or scatter plots to showcase trends.

```{r}
# Apply log1p (log(x + 1)) transformation to avoid log(0) issues
log_prices <- log1p(housing_data$`2`)

# Check the range of log-transformed prices
range_log_prices <- range(log_prices, na.rm = TRUE)

plot(housing_data$`3`, log_prices, main="House price vs Number of Bedrooms (Log-transformed)", 
     xlab = "Bedrooms", ylab = "Log of House Price + 1", 
     pch=19, cex = 1.5, col = rgb(0.2, 0.5, 0.8, 0.6), 
     xlim = c(min(housing_data$`3`), max(housing_data$`3`)), 
     ylim = range_log_prices)
```

This visualization shows the relationship between number of bedrooms and house prices, scaled to log(x+1). We scaled to log in order to make the graph more easy to understand visually, and did log(x+1) to eliminate the issue of taking the log of zero values. There is not a strong relationship presented by this graph, but we can see that the most expensive houses are concentrated at around 3-6 bedrooms. Because we used transparent blue points, we can see that the areas where the data points are most concentrated appear as a darker blue as many points are layered on top of each other. There is less data of houses with less than 3 or more than 6 bedrooms, as we can see the points still appear transparent. The visualization also depicts outliers at the bottom of the graph. These outliers have a significantly lower price than the rest of the dataset.

## PART 3: Expanding Your Investment Knowledge

### Why would this dataset be useful?

This additional dataset consists of the all-transactions house price index for Washington, US from 1975 to present. House price index (HPI) is an economic indicator that measures the change in average residential property prices over time. It shows how much house prices are rising or falling in a given region, usually calculated by tracking repeat sales of the same properties over time. It is considered a key tool for understanding housing market trends and affordability. 

This dataset would be useful because our dataset consists of data from Washington, US only from May-July 2014, so it would be useful to compare the data with the HPI in Washington throughout the years to see whether housing prices in the state were relatively low or high at that time.

### How could it complement the data you are currently analyzing?

This dataset could complement our dataset by showing the housing price index in 2014. We could compare the change in HPI from Quarter 2 and Quarter 3 to our own dataset. Our data was recorded from the months of May to August, which covers both quarters 2 and 3, so this will help us analyze the and compare the summary statistics of our data to the HPI of Washington in 2014. 

We could also use this dataset to compare the prices in our set to the trends of housing prices over the years. The HPI data shows that average residential property prices have risen over time in Washington. It shows a decline following the 2008 recession, with the lowest peak being in 2012, then a gradual increase. The pricing data in the original set is from 2014, so the data falls in a year of recovery from the recession, which showcases a unique trend that is captured in our data. The HPI had not yet hit the pre-recession high of 498.06 from 2007. It hit this high again in 2016 and has continued to increase since then. 

### Provide a link to the additional dataset.

[Additional dataset](https://fred.stlouisfed.org/series/WASTHPI)
